---
title: "p8105_hw2_amz2148"
output: github_document
---

Homework 2

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE) #shows all code chunks
```

# Problem 0

```{r load_libraries}
library(tidyverse) #loads tidyverse package
library(readxl) #loads readxl package
```

# Problem 1

First we read and clean the data from the Mr. Trash Wheel sheet under the following specifications:

* specifying the sheet in the Excel file and to omit non-data entries (rows with notes / figures; columns containing notes) using arguments in read_excel
* using reasonable variable names
* omitting rows that do not include dumpster-specific data
* rounding the number of sports balls to the nearest integer and converts the result to an integer variable (using as.integer)
* adding an additional variable to mark source of data (Mr. vs. Professor sheet)

```{r mr}
mr_trashwheel = 
  read_excel("data/Trash Wheel Collection Data.xlsx", range = "Mr. Trash Wheel!A2:N549") %>% #reads in only desired section and sheet from excel document
  janitor::clean_names()  %>% #cleans variable names
  drop_na(dumpster) %>% #removes rows with missing values for "dumpster" variable
  mutate(sports_balls = as.integer(sports_balls)) %>% #makes "sports_balls" variable an integer
  mutate(machine = "mister") %>% #creates new variable to identify sheet/source of data 
  mutate(year = as.double(year)) #converts character year variable to double
```

Then, we import, clean, and organize the data from the Professor Trash Wheel sheet.

```{r professor}
prof_trashwheel = 
  read_excel("data/Trash Wheel Collection Data.xlsx", range = "Professor Trash Wheel!A2:M96") %>% #reads in only desired section and sheet from excel document
  janitor::clean_names()  %>% #cleans variable names
  drop_na(dumpster) %>% #removes rows with missing values for "dumpster" variable
  mutate(machine = "professor") #creates new variable to identify sheet/source of data 
```

Finally, we combine the Mr. Trash Wheel and Professor Trash Wheel datasets to produce a single tidy dataset.

```{r bind}
trashwheel_tidy = 
  bind_rows(mr_trashwheel, prof_trashwheel) %>% #combines two datasets into one
  janitor::clean_names() %>% #cleans variable names
  select(machine, dumpster, everything()) %>% #reorders columns
  arrange (machine, dumpster) #sorts data by machine name and dumpster number
```

In the resulting dataset, there are ``r nrow(trashwheel_tidy)`` rows (observations) of which ``r nrow(subset(trashwheel_tidy,machine == "mister"))`` observations are from the Mr. Trashwheel dataset and ``r nrow(subset(trashwheel_tidy,machine == "professor"))`` observations are from the Professor Trashwheel dataset. The combined dataset has ``r ncol(trashwheel_tidy)`` variables of which one is new: ``machine`` which identifies which dataset the observation is from. Other important variables include ``date``, ``dumpster``, ``weight_tons``, ``volume_cubic_yards``, and ``homes_powered``. The total weight of trash collected by Professor Trashwheel was ``r sum(trashwheel_tidy[trashwheel_tidy$machine == "professor",]$weight_tons)`` tons. The total number of sports balls collected by Mr. Trash Wheel in 2020 was ``r sum(trashwheel_tidy[trashwheel_tidy$machine == "mister" & trashwheel_tidy$year == "2020",]$sports_balls)`` balls.


# Problem 2

First, we clean the data from the pols-month.csv file. We use separate() to break up the variable mon into integer variables year, month, and day; replace month number with month name; create a president variable taking values gop and dem, and remove prez_dem and prez_gop; and remove the day variable.

```{r pols_month}
pols_month = 
  read_csv(file = "data/fivethirtyeight_datasets/pols-month.csv") %>% 
  separate(mon, into = c("year", "month", "day"), sep = '-') %>%
  mutate(month = as.numeric(month), year = as.numeric(year), day = as.numeric(day)) %>%
  mutate(month = month.abb[month]) %>%
  mutate(month = factor(month, levels = month.abb)) %>%
  mutate(president = case_when(prez_dem == 1 ~ "dem",
    prez_dem != 1 ~ "gop")) %>%
  select(-prez_dem, -prez_gop, -day) %>%
  arrange(year, month)
```

Second, we clean the data in snp.csv using a similar process. For consistency across datasets, we arrange the data according to year and month, and organize so that year and month are the leading columns.




Third, tidy the unemployment data so that it can be merged with the previous datasets. This process will involve switching from “wide” to “long” format; ensuring that key variables have the same name; and ensuring that key variables take the same values.




Join the datasets by merging snp into pols, and merging unemployment into the result.




Write a short paragraph about these datasets. Explain briefly what each dataset contained, and describe the resulting dataset (e.g. give the dimension, range of years, and names of key variables).


