---
title: "p8105_hw2_amz2148"
output: github_document
---

Homework 2

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE) #shows all code chunks
```

# Problem 0

```{r load_libraries}
library(tidyverse) #loads tidyverse package
library(readxl) #loads readxl package
```

# Problem 1

First we read and clean the data from the Mr. Trash Wheel sheet:

* specifying the sheet in the Excel file and to omit non-data entries (rows with notes / figures; columns containing notes) using arguments in read_excel
* using reasonable variable names
* omitting rows that do not include dumpster-specific data
* rounding the number of sports balls to the nearest integer and converts the result to an integer variable (using as.integer)
* adding an additional variable to mark source of data (Mr. vs. Professor sheet)

```{r mr}
mr_trashwheel = 
  read_excel("data/Trash Wheel Collection Data.xlsx", range = "Mr. Trash Wheel!A2:N549") %>% #reads in only desired section and sheet from excel document
  janitor::clean_names()  %>% #cleans variable names
  drop_na(dumpster) %>% #removes rows with missing values for "dumpster" variable
  mutate(sports_balls = as.integer(sports_balls)) %>% #makes "sports_balls" variable an integer
  mutate(machine = "mister") %>% #creates new variable to identify sheet/source of data 
  mutate(year = as.double(year)) #converts character year variable to double
```

Then, we import, clean, and organize the data from the Professor Trash Wheel sheet.

```{r professor}
prof_trashwheel = 
  read_excel("data/Trash Wheel Collection Data.xlsx", range = "Professor Trash Wheel!A2:M96") %>% #reads in only desired section and sheet from excel document
  janitor::clean_names()  %>% #cleans variable names
  drop_na(dumpster) %>% #removes rows with missing values for "dumpster" variable
  mutate(machine = "professor") #creates new variable to identify sheet/source of data 
```

Finally, we combine the Mr. Trash Wheel and Professor Trash Wheel datasets to produce a single tidy dataset.

```{r bind}
trashwheel_tidy = 
  bind_rows(mr_trashwheel, prof_trashwheel) %>% #combines two datasets into one
  janitor::clean_names() %>% #cleans variable names
  select(machine, dumpster, everything()) %>% #reorders columns
  arrange (machine, dumpster) #sorts data by machine name and dumpster number
```

In the resulting dataset, there are ``r nrow(trashwheel_tidy)`` rows (observations) of which ``r nrow(subset(trashwheel_tidy,machine == "mister"))`` observations are from the Mr. Trashwheel dataset and ``r nrow(subset(trashwheel_tidy,machine == "professor"))`` observations are from the Professor Trashwheel dataset. The combined dataset has ``r ncol(trashwheel_tidy)`` variables of which one is new: ``machine`` which identifies which dataset the observation is from. Other important variables include ``date``, ``dumpster``, ``weight_tons``, ``volume_cubic_yards``, and ``homes_powered``. The total weight of trash collected by Professor Trashwheel was ``r sum(trashwheel_tidy[trashwheel_tidy$machine == "professor",]$weight_tons)`` tons. The total number of sports balls collected by Mr. Trash Wheel in 2020 was ``r sum(trashwheel_tidy[trashwheel_tidy$machine == "mister" & trashwheel_tidy$year == "2020",]$sports_balls)`` balls.


Note the number of observations in the resulting dataset and give examples of key variables. For available data, what was the total weight of trash collected by Professor Trash Wheel? What was the total number of sports balls collected by Mr. Trash Wheel in 2020?





# Problem 2




